---
title: "Facts Table Report"
author: Preethi Rajesh Yennemadi, Jazmyn Harris
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

# Setup
```{r setup, echo = T, eval=T, warning=FALSE, author = "Jazmyn Harris"}
library(DBI)
library(RMySQL)


DB_CERT <- <CERTIFICATE>

mart <- dbConnect(MySQL(),
                  user = "avnadmin",
                  password = <password>,
                  dbname = "defaultdb",
                  host = "practicum-i-cs5200-harrisj-jazmynrh12-9958.b.aivencloud.com",
                  port = 19929,
                  sslmode = "require",
                  sslcert = DB_CERT)
```

# Expense Analysis Report

## Intro

Heracletes Advisors PC is interested in keeping track of the business expenses that they accrued in order to bill clients. Expenses are accumulated from employee transactions that take place during travel for a given project. This expense reporting system allows employees to easily upload the transactions that they have garnered for a given credit card merchant, month, and year. The report system will extract information to produce different expense facts. This report will demonstrate an expense analysis from this reporting system.

\pagebreak
## Expense Facts Report

*GENERAL TABLE*

```{r efGeneral, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT *
FROM ExpenseFacts f
LIMIT 5"

dbGetQuery(mart, sql)
```

_Question 1: What are the total amount and average of each transaction spent (in USD) for each client in all time? (sorted by ClientName)_

```{r expenseFacts, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
    SELECT 
    cp.ClientName,
    SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD,
    AVG(f.AverageAmount * cu.USExchangeRate) AS AvgTransactionUSD
    FROM ExpenseFacts f
    JOIN ClientProjectDim cp ON f.ClientProjectDimID = cp.ClientProjectDimID
    JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
    GROUP BY cp.ClientName
    ORDER BY TotalSpentUSD DESC
    LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
### Time-Based

_Question 2: What is the trend of the total and average expense over time (by year and Month?_

```{r timeBased2, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT 
    f.Year,
    d.Month,
    SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD,
    AVG(f.AverageAmount * cu.USExchangeRate) AS AvgTransactionUSD
FROM ExpenseFacts f
JOIN DateDim d ON f.DateDimID = d.DateDimID
JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
GROUP BY f.Year, d.Month
ORDER BY f.Year, d.Month
"

df <- dbGetQuery(mart, sql)
df
```


*GRAPH*
```{r timeBased2_plot, echo=T, eval=T, warning=F, author="Jazmyn Harris"}
# Create a time label (Year-M)
df$Time <- paste(df$Year, "M", df$Month, sep="")
df$Time <- factor(df$Time, levels = df$Time)

# Expand right margin
par(mar = c(5, 4, 4, 10), xpd = TRUE)

# Plot Total and Average on same graph
plot(
  x = seq_along(df$Time),
  y = df$TotalSpentUSD,
  type = "o",
  xaxt = "n",
  xlab = "Time (Year-Month)",
  ylab = "USD Amount",
  main = "Expense Trends Over Time"
)

# Add x-axis labels
axis(1, at = seq_along(df$Time), labels = df$Time, las = 2, cex.axis = 0.7)

# Add Average line
lines(
  x = seq_along(df$Time),
  y = df$AvgTransactionUSD,
  type = "o",
  lty = 2
)

# Add legend
legend(
  "top",
  inset = c(-0.25, 0),
  legend = c("Total Spent (USD)", "Avg Transaction (USD)"),
  lty = c(1, 2),
  pch = 1,
  bty = "n"
)
```

\pagebreak
_Question 3: Which quarter has the highest spending overall across all clients?_
```{r timeBased3, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT 
    f.Year,
    f.Quarter,
    SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD
FROM ExpenseFacts f
JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
GROUP BY f.Year, f.Quarter
ORDER BY TotalSpentUSD DESC
LIMIT 1"

dbGetQuery(mart, sql)
```

\pagebreak
### Client-Based

_Question 4: Which clients/projects have the highest total spending across all time?_
```{r efFacts4, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT 
    cp.ClientName,
    cp.ProjectName,
    SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD
FROM ExpenseFacts f
JOIN ClientProjectDim cp ON f.ClientProjectDimID = cp.ClientProjectDimID
JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
GROUP BY cp.ClientName, cp.ProjectName
ORDER BY TotalSpentUSD DESC
LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
### Category-Based


_Question 5: Which expense categories have the highest total spent?_

```{r efFacts5, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT 
    c.CategoryName,
    SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD
FROM ExpenseFacts f
JOIN CategoryDim c ON f.CategoryDimID = c.CategoryDimID
JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
GROUP BY c.CategoryName
ORDER BY TotalSpentUSD DESC"

dbGetQuery(mart, sql)
```


_Question 6: What categories dominate spending for each client?_
```{r efFacts6, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT *
FROM (
    SELECT 
        cp.ClientName,
        c.CategoryName,
        SUM(f.TotalAmount * cu.USExchangeRate) AS TotalSpentUSD,
        RANK() OVER (
            PARTITION BY cp.ClientName 
            ORDER BY SUM(f.TotalAmount * cu.USExchangeRate) DESC
        ) AS rnk
    FROM ExpenseFacts f
    JOIN ClientProjectDim cp ON f.ClientProjectDimID = cp.ClientProjectDimID
    JOIN CategoryDim c ON f.CategoryDimID = c.CategoryDimID
    JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
    GROUP BY cp.ClientName, c.CategoryName
) ranked
WHERE rnk = 1;"

dbGetQuery(mart, sql)
```

\pagebreak
### Currency-Based


_Question 7: Which currency has the most transactions?_ 
```{r efFacts7, echo=T, eval=T, warning=F, author = "Jazmyn Harris"}
sql <- "
SELECT 
    cu.CurrencyName,
    COUNT(*) AS TransactionCount
FROM ExpenseFacts f
JOIN CurrencyDim cu ON f.CurrencyDimID = cu.CurrencyDimID
GROUP BY cu.CurrencyName
ORDER BY TransactionCount DESC
LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
## Billable Facts Report

*GENERAL TABLE*

```{r bfGeneral, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT *
FROM BillableFacts
LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
_Question 1: What is the total billable vs non-billable amount over time?_

```{r billablefact1, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT 
    f.Year,
    f.Quarter,
    SUM(f.BillableAmountUSD) AS TotalBillableUSD,
    SUM(f.NonBillableAmountUSD) AS TotalNonBillableUSD
FROM BillableFacts f
GROUP BY f.Year, f.Quarter
ORDER BY f.Year, f.Quarter"

df <- dbGetQuery(mart, sql)
df
```

```{r billablefact1_plot, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
# Create time label
df$Time <- paste(df$Year, "Q", df$Quarter, sep="")
df$Time <- factor(df$Time, levels = df$Time)

# Expand right margin for legend
par(mar = c(5, 4, 4, 10), xpd = TRUE)

# Plot Billable
plot(
  x = seq_along(df$Time),
  y = df$TotalBillableUSD,
  type = "o",
  xaxt = "n",
  xlab = "Time (Year-Quarter)",
  ylab = "USD Amount",
  main = "Billable vs Non-Billable Amount Over Time"
)

# X-axis labels
axis(1, at = seq_along(df$Time), labels = df$Time, las = 2, cex.axis = 0.7)

# Add Non-Billable line
lines(
  x = seq_along(df$Time),
  y = df$TotalNonBillableUSD,
  type = "o",
  lty = 2
)

# Legend on the right
legend(
  "topright",
  inset = c(-0.25, 0),
  legend = c("Billable Amount (USD)", "Non-Billable Amount (USD)"),
  lty = c(1, 2),
  pch = 1,
  bty = "n"
)
```

\pagebreak
_Question 2: Which quarter has the highest billable revenue?_

```{r billablefact2, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT 
    f.Year,
    f.Quarter,
    SUM(f.BillableAmountUSD) AS TotalBillableUSD
FROM BillableFacts f
GROUP BY f.Year, f.Quarter
ORDER BY TotalBillableUSD DESC
LIMIT 1"

dbGetQuery(mart, sql) 
```


_Question 3: Which employees generate the highest billable revenues?_

```{r billablefact3, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT 
    e.EmployeeName,
    SUM(f.BillableAmountUSD) AS TotalBillableUSD
FROM BillableFacts f
JOIN EmployeeDim e 
    ON f.EmployeeDimID = e.EmployeeDimID
GROUP BY e.EmployeeName
ORDER BY TotalBillableUSD DESC;"

dbGetQuery(mart, sql) 
```

\pagebreak
## BudgetVsActual Facts Report

*GENERAL TABLE*

```{r bvaGeneral, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT *
FROM BudgetVsActualFacts
LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
_Question 1: Which projects have the worst budget accuracy?_

```{r bvaFacts1, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
sql <- "
SELECT 
    cp.ProjectName,
    SUM(f.ProjectBudget) AS TotalBudget,
    SUM(f.ActualSpendUSD) AS TotalActual,
    SUM(f.Variance) AS TotalVariance,
    (SUM(f.Variance) / NULLIF(SUM(f.ProjectBudget), 0)) * 100 AS VariancePercent
FROM BudgetVsActualFacts f
JOIN ClientProjectDim cp 
    ON f.ClientProjectDimID = cp.ClientProjectDimID
GROUP BY cp.ProjectName
ORDER BY ABS(VariancePercent) DESC
LIMIT 5"

dbGetQuery(mart, sql)
```

\pagebreak
_Question 2: What is the variance trend year-over-year?_

```{r bvaFacts2, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}

sql <- "
SELECT 
    f.Year,
    SUM(f.ProjectBudget) AS TotalBudget,
    SUM(f.ActualSpendUSD) AS TotalActual,
    SUM(f.Variance) AS TotalVariance,
    (SUM(f.Variance) / NULLIF(SUM(f.ProjectBudget), 0)) * 100 AS VariancePercent
FROM BudgetVsActualFacts f
GROUP BY f.Year
ORDER BY f.Year
LIMIT 10"

df <-dbGetQuery(mart, sql)
df
```


*GRAPH*
```{r bvaFacts2_plot, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}
# preventing scientific notation use
options(scipen = 999)

# Time label
df$Time <- as.character(df$Year)
df$Time <- factor(df$Time, levels = df$Time)

# Expand margin for legend
par(mar = c(5, 4, 4, 10), xpd = TRUE)

# Plot Budget
plot(
  x = seq_along(df$Time),
  y = df$TotalBudget,
  type = "o",
  xaxt = "n",
  yaxt = "n",       
  xlab = "Year",
  ylab = "USD Amount",
  main = "Budget vs Actual Spending Over Time"
)

# Y-axis
axis(2, at = axTicks(2), labels = format(axTicks(2), scientific = FALSE), cex.axis = 0.7)

# X-axis labels
axis(1, at = seq_along(df$Time), labels = df$Time, las = 2)

# Add Actual line
lines(
  x = seq_along(df$Time),
  y = df$TotalActual,
  type = "o",
  lty = 2
)

# Legend on the side
legend(
  "topright",
  inset = c(-0.25, 0),
  legend = c("Budget", "Actual Spend"),
  lty = c(1, 2),
  pch = 1,
  bty = "n"
)
```

\pagebreak
_Question 3: What is the variance trend year and quarter?_

```{r bvaFacts3, echo=T, eval=T, warning=F, author="Preethi Yennemadi"}

sql <- "
SELECT 
    f.Year,
    f.Quarter,
    SUM(f.ProjectBudget) AS TotalBudget,
    SUM(f.ActualSpendUSD) AS TotalActual,
    SUM(f.Variance) AS TotalVariance,
    (SUM(f.Variance) / NULLIF(SUM(f.ProjectBudget), 0)) * 100 AS VariancePercent
FROM BudgetVsActualFacts f
GROUP BY f.Year, f.Quarter
ORDER BY f.Year, f.Quarter
LIMIT 5"

dbGetQuery(mart, sql)

```


```{r disconnect, echo=T, eval=T, warning=F, author="Jazmyn Harris"}
dbDisconnect(mart)
```

